### 2016-02-03 inputs/discussion.tex
update in results, discussion

\subsection{Minor Allele Frequency and data (un)balancedness}
\label{sec:maf}
GBS data pose a greater challenge than SNP-chip data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they usually contain. On our data sets we found missing rates varying from 53\% to 67\%. With larger amounts of missing data the complexity of the imputation problem tends to increase. Moreover, with such a high level of missingness, even imputation errors as low as 1\% may have negative impact on successive analysis, amplifying the importance of finding the most accurate imputation method.\\
The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points. Classification problems are known to be harder when data are unbalanced, i.e. the classes appear at different frequencies in the datasets, with typically one over-represented class
%problem with url
(see~\cite{kotsiantis_handling_2006}
and~\cite{sun_classification_2009} for a review). In SNP genotype imputation, data balancedness is directly related to the minor allele frequency (MAF). In the classification of unbalanced observations, it is important to look not only at the total classification accuracy, but also at the per-class accuracy. The total classification accuracy may be misleading, being ``dominated'' by the majority class
%problem with url
\cite{he_learning_2009}.
Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present on minority classes. 
The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans). Our per-class dissection of results allowed a deeper insight into the imputation process. Indeed, all imputation algorithms performed considerably better in the majority class rather than in the less frequent classes.
Some algorithms failed almost completely the imputation of the least common genotypes, while other algorithms proved to work well. In alfalfa, KNNI was easily the best imputation method in the heterozygous and minor homozygous classes, while SVDI and RFI performed only slightly better than MNI. In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes. MNI performed reasonably well in the alfalfa heterozygous class, with accuracy close to 0.6, contributed to its moderately good results (total accuracy around 0.8).

\subsection{Missing rate and the ``curse of dimensionality''}
\label{sec:missing_rate_curse_dimensionality}
In general, imputation methods were relatively robust to increasing missing rates. Only KNNI, and to a lesser extent RFI, showed slowly degrading performances at very high missing rates. The divergent response of the different imputation methods to missing rate became apparent only in the most challenging scenarios, when markers with up to 40-70\% missing genotypes were allowed in the data set.\\
KNNI's relatively pronounced drop in imputation accuracy with increasing missing rates can be interpreted in relation to the phenomenon known as the ``curse of dimensionality'' \cite{bellman_dynamic_1957,marimont_nearest_1979}: with the same amount of data, the increasing number of parameters - the ``dimensions'' of the problem - increases and complicates the identification of k neighbors which are close enough to the data point to be classified/imputed. 
On average, the side $l$ of the hypercube to include $k$ neighbors is a function of $k$, $n$ (sample size) and $p$ (number of parameters): $l = \left ( \frac{k}{n} \right )^{1/p}$. With $n$ constant, the hypercube in which the $k$ neighbors lie gets bigger as $p$ increases. This holds especially for local learning methods (e.g.,~K-Nearest Neighbors methods, local regression) which rely heavily on the the information content of the neighbourhood. RFI - which only partially relies on local structure of the data - suffered marginally from high missingness. SVD and MNI, which build on more general properties of the data, and Beagle - whose learning process is fundamentally different and specific for genotype imputation - were scarcely affected by missing rates.

\subsection{Imputation efficiency: differences in rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
%problems with this citation: url contains hashes
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: 
i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; \\
ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);\\
iii) alfalfa datasets did not contain population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulation. Population structure is implicitly used by algorithms as Beagle and KNNI that create clusters of similar genotypes. Thus, rice dataset population structure facilitated the inputation process;\\
iv) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.

>>>>

\subsection{Minor Allele Frequency and data (un)balancedness}
\label{sec:maf}
GBS data pose a greater challenge than SNP-array data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they contain. On our data sets we found missing rates varied from 53\% to 67\%. With larger amounts of missing data the complexity of the imputation problem increases. Imputation error can have a negative impact on successive analyses (see for instance \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points. Classification problems are known to be harder when data are unbalanced, i.e. the classes appear at different frequencies in the datasets, with typically one over-represented class
(see~\cite{kotsiantis_handling_2006}
and~\cite{sun_classification_2009} for a review). In SNP genotype imputation, data balancedness is directly related to the minor allele frequency (MAF). In the classification of unbalanced observations, it is important to look not only at the total classification accuracy, but also at the per-class accuracy. The total classification accuracy may be misleading, being ``dominated'' by the majority class
\cite{he_learning_2009}.
Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present on minority classes. 
The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans). Our per-class dissection of results allowed a deeper insight into the imputation process. Indeed, all imputation algorithms performed considerably better in the majority class rather than in the less frequent classes.
In alfalfa, KNNI was easily the best imputation method in the heterozygous and minor homozygous classes, while SVDI and RFI performed only slightly better than MNI. In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes. MNI performed reasonably well in the alfalfa heterozygous class, with accuracy close 60\%, contributed to its moderately good results (total accuracy around 80\%).

\subsection{Missing rate and the ``curse of dimensionality''}
\label{sec:missing_rate_curse_dimensionality}
In general, imputation methods were relatively robust to increasing missing rates. Only KNNI, and to a lesser extent RFI, showed slowly degrading performances at very high missing rates. The divergent response of the different imputation methods to missing rate became apparent only in the most challenging scenarios, when markers with up to 40-70\% missing genotypes were allowed in the data set.\\
KNNI's decrease in imputation accuracy with increasing missing rates can be interpreted in relation to the phenomenon known as the ``curse of dimensionality'' \cite{bellman_dynamic_1957,marimont_nearest_1979}: with the same amount of data, the increasing number of parameters - the ``dimensions'' of the problem - increases and complicates the identification of k neighbors which are close enough to the data point to be classified/imputed. 
On average, the side $l$ of the hypercube to include $k$ neighbors is a function of $k$, $n$ (sample size) and $p$ (number of parameters): $l = \left ( \frac{k}{n} \right )^{1/p}$. With $n$ constant, the hypercube in which the $k$ neighbors lie gets bigger as $p$ increases. This holds especially for local learning methods (e.g.,~K-Nearest Neighbors methods, local regression) which rely heavily on the the information content of the neighbourhood. RFI - which only partially relies on local structure of the data - suffered marginally from high missingness. SVD and MNI, which build on more general properties of the data, and Beagle - whose learning process is fundamentally different and specific for genotype imputation - were scarcely affected by missing rates.

\subsection{Imputation efficiency: differences in rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: \\
i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; \\
ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);\\
iii) alfalfa datasets did not contain population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulation. Population structure is implicitly used by algorithms as Beagle and KNNI that create clusters of similar genotypes. Thus, rice dataset population structure facilitated the inputation process;\\
iv) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.\\
The most dramatic difference between data sets was found in FILLIN performances. The algorithm costantly produced 100\% uninputed results in alfalfa, while resulted in average performances in rice (albeit inferior to the other algorithms). This behaviour is probably due to FILLIN being designed to target inbred lines (as the name itself indicates).

### 2016-02-03 inputs/results.tex
update in results, discussion

\subsection{Genotypes}
\label{sec:genotypes}  
GBS detected 32\,706, 40\,734 and 166\,418 SNP loci in Alfalfa-PV, Alfalfa-Med and Rice, respectively (Table~\ref{tab:descriptive_statistics}). The overall missing rate was $66.6\%$, $59.6\%$ and $53.4\%$ in the three datasets. Supplementary Figure S1 shows the density distribution of missing rate per markers in the complete datasets. The amount of missing marker genotypes varied with the threshold of maximum per-marker missing-rate allowed in the data (10\%, 20\%, 40\% and 70\%) and the proportion of artificial missing genotypes that were introduced (1\%, 5\%, 10\% and 20\%). Missing data point counts ranged from a minimum of 3\,878 in Alfalfa-PV with 10\% allowed and 1\% artificial missing genotypes, to a maximum of 5\,580\,616 in Rice with 70\% allowed and 20\% artificial missing genotypes. Over all datasets and allowed/artificial missing genotypes thresholds, the amount of missing data points to be imputed averaged 249\,405.\\
Minor Allele Frequency (MAF) was 0.172, 0.170 and 0.140 in Alfalfa-PV, Alfalfa-Med and Rice, respectively. The three datasets were therefore unbalanced with respect to genotype classes: Rice had the lowest MAF, while Alfalfa-PV and Alfalfa-Med had the smallest minor classes (minor homozygotes 3.5\% and 3.2\% of the total genotypes). Figure~\ref{fig:genotype_classes} reports the proportion of genotypes in each class in the complete datasets (100\%) and as a function of the maximum allowed missing rate per marker. The relative proportion of genotype classes remained approximately stable in rice. In contrast, the heterozygous class in alfalfa tended to become smaller with increasing proportion of allowed genotypes. This reflected the different GBS calling criteria for homozygous and heterozygous SNP: heterozygous SNPs required fewer overlapping reads to be called, while the implemented quality filters on alfalfa required a larger number of reads to call a homozygous loci. The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.

\textcolor{red}{Some numbers on alignment on reference genome.}

@@

\subsection{Imputation accuracy}
\label{sec:imputation_accuracy}

\textcolor{red}{
\begin{itemize}
  \item FILLIN on Rice: explain uninputed fraction
  \item FILLIN on Alfalfa: FAIL
  \item all rice chromosome gave the same results, as in Suppl figures
\end{itemize}
}
In rice, the average imputation accuracy over all genotypic classes was close to 100\% with most methods and still above 90\% at maximum proportions of allowed (70\%) and artificial (20\%) missing genotypes (Figure~\ref{fig:accuracy_rice}). The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 84\% (Figure~\ref{fig:accuracy_alfalfa}). The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype class (averaging 90.8\% and 99.3\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging 4.1\% and 73.4\% in alfalfa and rice, respectively). Alfalfa datasets included heterozygous SNP loci, accounting for an average of 30\% of individuals per locus (Table~\ref{tab:descriptive_statistics}). Heterozygotes had an intermediate imputation accuracy ranging from 0\% with MNI to 70\% with KNNI. Results for each individual alfalfa population, the 12 individual rice chromosomes, and the whole rice dataset are similar to those discussed here (see Supplementary Figures S2 to S16). \\
Missing data did not affect the imputation accuracy substantially, with most algorithms showing a flat or almost flat response to increased missing rate. KNNI showed a decreasing imputation accuracy with increasing missing data, both in alfalfa and rice.\\
In rice, Beagle outperformed all other imputation methods, making efficient use of marker position (Figure~\ref{fig:accuracy_rice}). When markers were randomly reshuffled (thus offsetting position information), though, general imputation methods (other than MNI) showed a higher imputation accuracy than Beagle. In general we observed for all thresholds and in both genotype classes the following ranking of methods for imputation accuracy: RFI $>$ KNNI $>$ SVDI.
In particular, for 70\% allowed missing rate datasets, the imputation accuracy of RFI was still greater than 90\%, while it dropped towards 75\% for KNNI and SVDI. As expected, MNI had an imputation accuracy of 100\% in the major homozygous class, but dropped to 0\% in the minor homozygous class.\\
In alfalfa KNNI showed the highest imputation accuracy over all classes and in the heterozygous and minor homozygous genotype classes (Figure~\ref{fig:accuracy_alfalfa}). RFI and SVDI were second ranking. They did not differ much from MNI in the overall accuracy and in the major homozygous class, but gave higher imputation accuracy in the heterozygous and minor homozygous classes. Beagle was also applied to alfalfa, where the relative position of markers in the genome is not known. The overall imputation accuracy with Beagle was about 67\%, with accuracies of almost 100\% in the major homozygous class but practically 0\% in the heterozygous and minor homozygous classes.

>>>>

\subsection{Genotypes}
\label{sec:genotypes}  
GBS detected 32\,706, 40\,734 and 166\,418 SNP loci in Alfalfa-PV, Alfalfa-Med and Rice, respectively (Table~\ref{tab:descriptive_statistics}). The overall missing rate was $66.6\%$, $59.6\%$ and $53.4\%$ in the three datasets. Supplementary Figure S1 shows the density distribution of missing rate per markers in the complete datasets. The amount of missing marker genotypes varied with the threshold of maximum per-marker missing-rate allowed in the data (10\%, 20\%, 40\% and 70\%) and the proportion of artificial missing genotypes that were introduced (1\%, 5\%, 10\% and 20\%). Missing data point counts ranged from a minimum of 3\,878 in Alfalfa-PV with 10\% allowed and 1\% artificial missing genotypes, to a maximum of 5\,580\,616 in Rice with 70\% allowed and 20\% artificial missing genotypes. Over all datasets and allowed/artificial missing genotypes thresholds, the amount of missing data points to be imputed averaged 249\,405.\\
Minor Allele Frequency (MAF) was 0.172, 0.170 and 0.140 in Alfalfa-PV, Alfalfa-Med and Rice, respectively. The three datasets were therefore unbalanced with respect to genotype classes: Rice had the lowest MAF, while Alfalfa-PV and Alfalfa-Med had the smallest minor classes (minor homozygotes 3.5\% and 3.2\% of the total genotypes). Figure~\ref{fig:genotype_classes} reports the proportion of genotypes in each class in the complete datasets (100\%) and as a function of the maximum allowed missing rate per marker. The relative proportion of genotype classes remained approximately stable in rice. In contrast, the heterozygous class in alfalfa tended to become smaller with increasing proportion of allowed genotypes. This reflected the different GBS calling criteria for homozygous and heterozygous SNP: heterozygous SNPs required fewer overlapping reads to be called, while the implemented quality filters on alfalfa required a larger number of reads to call a homozygous loci. The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.
\textcolor{red}{
Alignment on reference genome assigned a position to 88.66\%, xxx\%, and xxx\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively.
}

@@

\subsection{Imputation accuracy}
\label{sec:imputation_accuracy}  
\textcolor{red}{
\begin{itemize}
  \item all rice chromosome gave the same results, as in Suppl figures
  \item all numbers in this section need to be checked
  \item what about supplementary figures? still valid?
\end{itemize}
}
In rice, the average imputation accuracy over all genotypic classes was above 80\% for all methods. No significant difference was found between chromosome, thus results are presented averaged (Figure~\ref{fig:accuracy_rice}). \\
The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 84\% (Figure~\ref{fig:accuracy_alfalfa}). Again, since no significant difference was found between the two dataset, presented results are averaged (Figure~\ref{fig:accuracy_alfalfa}).

The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype class (averaging 90.8\% and 99.3\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging 4.1\% and 73.4\% in alfalfa and rice, respectively).

Alfalfa datasets included heterozygous SNP loci, accounting for an average of 30\% of individuals per locus (Table~\ref{tab:descriptive_statistics}). Heterozygotes had an intermediate imputation accuracy ranging from 0\% with MNI to 70\% with KNNI. Results for each individual alfalfa population, the 12 individual rice chromosomes, and the whole rice dataset are similar to those discussed here (see Supplementary Figures S2 to S16).

Missing data did not affect the imputation accuracy substantially, with most algorithms showing a flat or almost flat response to increased missing rate. KNNI showed a decreasing imputation accuracy with increasing missing data, both in alfalfa and rice.

In rice, Beagle outperformed all other imputation methods, making efficient use of marker position (Figure~\ref{fig:accuracy_rice}). When markers were randomly reshuffled (thus offsetting position information), though, general imputation methods (other than MNI) showed a higher imputation accuracy than Beagle. The loss of performances on reshuffled markers is exacerbated on alfalfa datasets, and in particular on heterozygous and minor homozygous classes.

Similarly, FILLIN resulted in an average drop of XXX\% when marker positions are shuffled. FILLIN performances were in general lower that other methods due to the fraction of unimputed genotypes, here accounted as wrong imputation. FILLIN did not impute an average of XXX\% missing data in Rice datasets and practically the entirety of Alfalfa dataset. Even trying to relax FILLIN constraint parameters, it was impossible to obtain any sensible result.

In general we observed for all thresholds and in both genotype classes the following ranking of methods for imputation accuracy: RFI $>$ KNNI $>$ SVDI.
In Rice, for 70\% allowed missing rate datasets, the imputation accuracy of RFI was still greater than 90\%, while it dropped towards 75\% for KNNI and SVDI. As expected, MNI had an imputation accuracy of 100\% in the major homozygous class, but dropped to 0\% in the minor homozygous class.\\
In alfalfa KNNI showed the highest imputation accuracy over all classes and in the heterozygous and minor homozygous genotype classes (Figure~\ref{fig:accuracy_alfalfa}). RFI and SVDI were second ranking. They did not differ much from MNI in the overall accuracy and in the major homozygous class, but gave higher imputation accuracy in the heterozygous and minor homozygous classes.
